### 2015-05-13 whoop.tex

We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.

@@

\medskip\noindent\textbf{Lockset Instrumentation }
%
Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibility that $S$ and $T$ are equal. We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is an instance of $T$, in the presence of arbitrarily many further threads that may execute concurrently.

@@

\ADComment{Renaming must talk about basic block exit conditions} Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.  Shared variables and mutexes are \emph{not} renamed, because these are shared across templates.

@@

Initially, lockset instrumentation introduces sets that track: the shared variables that have been read from and written to; the mutexes that have been consistently used to protect each variable; and the mutexes that are currently held by a thread.  To this end, when instrumenting template $T$ with respect to symbol $i$, we add the sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$. For each shared variable $s \in \sharedvars$ we add a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$. Finally, we add a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread.

The statements of each procedure in $\instrument{T}{i}$ that manipulate shared variables and mutexes are then instrumented to take account of the above sets.  This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.  A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).  Notice that in the instrumented form, the shared variable $s$ is no longer assigned to.  We discuss the reasons for this when we describe \emph{shared state abstraction} below.  A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command applied to the receiving private variable $x$; again this relates to shared state abstraction and is discussed below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.

>>>>

We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

\medskip\noindent\textbf{Lockset Instrumentation }
%
For thread templates $S$ and $T$ (including the possibility that $S$ and $T$ are equal) we wish to check whether it is possible for a thread executing $S$ to race with a thread executing $T$, in the presence of arbitrarily many further concurrently executing threads.

@@

Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $S$ and $T$ are the same.

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.

The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.  This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.  A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).  A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.
